### 2016-11-08 Manuscript/SW_early_life_telomeres.Rmd
Eds

Longitudinal data showed that early-life RTL was weakly, but significantly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,4])`; Fig. 4A). $\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(dd3,Agemonths == 1)$DeltaRTL)$statistic)`; P `r round3(t.test(subset(dd3,Agemonths == 1)$DeltaRTL)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(dd3,Agemonths == 6)$DeltaRTL)$statistic)`; P = `r round2(t.test(subset(dd3,Agemonths == 6)$DeltaRTL)$p.value)`) or subadults (t = `r round2(t.test(subset(dd3,Agemonths == 10)$DeltaRTL)$statistic)`; P = `r round2(t.test(subset(dd3,Agemonths == 10)$DeltaRTL)$p.value)`), suggesting that telomere shortening was highest shortly after the nestling phase, after which both telomere shortening and telomere lengthening were observed (Fig. 4B).

To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details). We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 4C). Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`). We found no evidence that the likelihood of telomere lengthening was related to the follow up time between sampling events (GLM: t = `r round2(summary(glm(DeltaRTLF~TimeDiff,data=dd3,family = 'binomial'))$coefficients[2,3])`; P = `r round2(summary(glm(DeltaRTLF~TimeDiff,data=dd3,family = 'binomial'))$coefficients[2,4])`), and considering those samples that did show increases in $\Delta$RTL, the degree of increase was not related to follow up time (LM: t = `r round2(summary(lm(DeltaRTL~TimeDiff,data=subset(dd3,DeltaRTL>0)))$coefficients[2,3])`; P = `r round2(summary(lm(DeltaRTL~TimeDiff,data=subset(dd3,DeltaRTL>0)))$coefficients[2,4])`).

@@

summary(glmer(SurvivedNext1~DeltaRTL+LogAge+(1|LayYear),data=dd3,family = 'binomial'))

>>>>

Longitudinal data showed that early-life RTL was weakly, but significantly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,4])`; Fig. 1B). As with cross-sectional RTL, $\Delta$RTL decreased with age (R^2^ = `r round3(summary(lm(DeltaRTL~Agemonths,data = dd3))$r.squared)`; t = `r round2(summary(lm(DeltaRTL~Agemonths,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(DeltaRTL~Agemonths,data = dd3))$coefficients[2,4])`), suggesting that telomere shortening was highest shortly after the nestling phase, after which both telomere shortening and telomere lengthening were observed.

To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details). We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 1C). Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`). We found no evidence that the likelihood of telomere lengthening was related to the follow up time between sampling events (GLM: t = `r round2(summary(glm(DeltaRTLF~TimeDiff,data=dd3,family = 'binomial'))$coefficients[2,3])`; P = `r round2(summary(glm(DeltaRTLF~TimeDiff,data=dd3,family = 'binomial'))$coefficients[2,4])`), and considering those samples that did show increases in $\Delta$RTL, the degree of increase was not related to follow up time (LM: t = `r round2(summary(lm(DeltaRTL~TimeDiff,data=subset(dd3,DeltaRTL>0)))$coefficients[2,3])`; P = `r round2(summary(lm(DeltaRTL~TimeDiff,data=subset(dd3,DeltaRTL>0)))$coefficients[2,4])`).

@@

summary(glmer(SurvivedNext~RTL+LogAge+Sex+Tarsus+(1|LayYear),data=subset(dd14,Agemonths>30),family = 'binomial'))

summary(coxph(Surv(RemainingLife,Died)~RTL+LogAge,data=dd8))
